Variation matters: Expanding the scope of experimental archaeology

### 2024-07-25 manuscript/manuscript.Rmd
title and readme file update.

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: 
|
  ```{=latex}
  This paper presents the Perception-Process-Product ("Triple P") framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record. It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.

  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

Built upon this critique of *Homo economicus* as well as intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions. To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices. These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea. The second principle particularly allows researchers to develop research questions that are also meaningful to descendant communities through respectful conversation and collaboration [@montgomery2023]. The Triple P framework also **3)** adopts a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects. It is acknowledged that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@eren2016; @roe2009: 1266-1267]. Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions. This research orientation prioritizes the material science aspect over the social science aspect of experimental archaeology. Similarly, standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023]. In real-world knapping, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022].

On the other hand, less-controlled experiments, which have been traditionally known as naturalistic or actualistic experiments [see @conrad2023; @eren2024 for detailed terminological critiques], pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, a less-controlled experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Unlike strictly controlled experiments testing one variable of interest each time [@almaatouq2024], less-controlled experiments are designed to produce variation and their interactions. This feature is crucial and cannot be simply replaced by ethnographic records or ethnoarchaeology, because many paleolithic technological components do not have analogs in contemporary non-industrial societies [e.g., @arthur2018; @stout2002]. While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by less-controlled experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023]. Less-controlled experiments can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681], who proposed that the interaction between less-controlled and strictly controlled experiments "operates in a cyclical form of induction and deduction."

@@

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century [@flenniken1984; @coles1979; @johnson1978; @whittaker1994: 54-61]. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to generate ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree [-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artifacts. Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022; @milks2023] across different regions. Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023; @marshall2002; @douglass2020].

However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation. The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors. This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research. To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations.

@@

@@

Ethnographies revolving around experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S. hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel. However, ethnography has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. Within the broader context of burgeoning interest in mixed-method research in contemporary social science [@creswell2017], this also echoes the post-positivist turn in psychology in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].

Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences, social interactions, and cultural meanings associated with the experimental activities [@gowlland2019]. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage. Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely "an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen)." In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography becomes necessary.

## Multi-level data curation

@@

Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project. Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researchers; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.

@@

I thank Dietrich Stout for his helpful comments on earlier drafts of this article as well as Mark Moore and Shaya Jannati for inspiring discussions. This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."

# References

>>>>

---
title: "Variation matters: Expanding the scope of experimental archaeology"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: 
|
  ```{=latex}
  \textbf{Abstract:} This paper aims to expand the scope of experimental archaeology to emphasize multi-level variation and interactions across the levels of perception, actions, and outcomes. Such an approach, loosely formulated as the Perception-Process-Product ("Triple P") framework, offers a more grounded and richer explanation of the past archaeological record. It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects. Serving as a heuristic device, this alternative mode of knowledge production is highly flexible in nature, where each single component is detachable as dictated by individual research questions.

  \textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production

  \textbf{Abstracto:} El objetivo de este artículo es ampliar el alcance de la arqueología experimental para centrar la atención en la variación multinivel y las interacciones entre los rangos de percepción, acciones y resultados. Este enfoque, formulado vagamente como el marco Percepción-Proceso-Producto ("Triple P"), ofrece una explicación más fundamentada y rica del registro arqueológico del pasado. Consta de tres principios 1) reconocer la compensación inherente entre control y generalizabilidad en el diseño de la investigación experimental; 2) fomentar proyectos de colaboración que impliquen a participantes en la investigación de procedencias distintas no limitados a los tradicionales participantes, como aficionados y novatos; 3) adoptar un flujo de trabajo que normalice la recogida y conservación de datos etológicos y etnográficos en proyectos experimentales. Como dispositivo heurístico, este modo alternativo de producción de conocimientos es muy flexible por naturaleza, y cada uno de sus componentes es desechable o modificable en función de las preguntas de investigación individuales.

  \textbf{Palabras clave:} Arqueología experimental; Análisis etológico; Investigación etnográfica; Maldición del conocimiento; Producción colaborativa de conocimientos
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

@@

Built upon this critique of *Homo economicus* as well as the four strategies$^1$ of behavioral archaeology [@schiffer2010; @reid1975], here I propose the Triple P framework, which aims to **a)** amplify the expression of variation in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) by experiments in diverse contexts and **b)** better identify the complex interacting relationships across these three levels of variations in real-world conditions. To accomplish these two objectives, I advocate the following three principles as integral components of the Triple P framework, which requires **1)** acknowledging the inherent trade-off between control and generalizability in the experimental research design and **2)** encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices. These two principles are developed to advocate a pluralistic approach to the explanation of complex variation, which has received more attention from evolutionary anthropology [@antón2017] to cognitive science [@barrett2020], instead of treating the optimization-based research agenda as a panacea. The second principle particularly allows researchers to develop research questions that are also meaningful to descendant communities through respectful conversation and collaboration [@montgomery2023]. The Triple P framework also **3)** adopts a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects. It is acknowledged that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

The trade-off between causal inference (aka "internal validity") and generalization (aka "external validity") forms a central issue in experimental design across different disciplines [@degtiar2023; @roe2009: 1266-1267]$^2$. Even in fields known for their development of rigorous and well-controlled experimental methods such as cognitive psychology and neuroscience, researchers have started to use relatively naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiments due to the generalizability crisis, namely the prevailing mismatch between phenomenon of interest and measured variables in psychological science [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. In contrast, the past decades have witnessed experimental archaeology's growing research interests focusing on the robust inference of causal mechanisms while compromising generalizability in the explanation of material culture variation [@eren2016; @lin2018; @eren2024; @marreiros2020]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design can be insufficient in inferring how context-generic principles interact in a particular context as reflected in real-world conditions. This research orientation prioritizes the material science aspect over the social science aspect of experimental archaeology. Similarly, standardized artificial materials like bricks [@lombao2017] or foam blocks [@schillinger2016] have been used to standardize materials and/or reduce learning demands in experimental studies focusing on the transmission of lithic technologies, with implications for the generalizability of results [@liu2023]. In real-world knapping, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in cultural transmission and skill development [@proffitt2022].

On the other hand, less-controlled experiments, which have been traditionally known as naturalistic or actualistic experiments [see @conrad2023; @eren2024 for detailed terminological critiques], pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, a less-controlled experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Unlike strictly controlled experiments testing one variable of interest each time [@almaatouq2024], less-controlled experiments are designed to produce variation and their interactions. This feature is crucial and cannot be simply replaced by ethnographic records or ethnoarchaeology, because many paleolithic technological components do not have analogs in contemporary non-industrial societies [e.g., @arthur2018; @stout2002]. While uncontrolled variation has traditionally been viewed as highly problematic, statistical techniques for developing causal inference from observational data, of the kind produced by less-controlled experiments, have also been greatly boosted in epidemiology and economics in recent years [@cunningham2021; @hernan2023]. Despite the fact that one should not interpret any experiment as a direct representation of an actual past event [@eren2024], less-controlled experiments can serve a heuristic role in hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681], who proposed that the interaction between less-controlled and strictly controlled experiments "operates in a cyclical form of induction and deduction."

@@

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the nineteenth century [@johnson1978; @whittaker1994: 54-61]. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to generate ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree [-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] ethnography. Likewise, novices' lack of flintknapping expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts. Several researchers have also pointed out that literature-informed archaeologists sometimes get lost in reconstructing previous archaeologists’ reconstructions instead of searching for diverse solutions to better understand the actual archaeological phenomenon [@bell2014; @currie2022], which is another reason why we need the presence of hobbyists and novices in the community of experimental archaeology.

Emphasizing variation at its core, the Triple P conceptual framework recognizes that experimental archaeology can greatly benefit from diverse perspectives [@pargeter2023: 164] and thereby inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2023] of stone artifacts. Furthermore, the Triple P framework acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage [@arthur2024]. This emphasis on team efforts and inclusivity allows for a more complete understanding of the non-utilitarian or unexpected aspects of raw material procurement [@batalla2016] and selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022; @milks2023] across different regions. Through ethical collaborations with those knapping practitioners in non-industrial societies in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@montgomery2023; @marshall2002; @douglass2020].

However, the facilitation of large-scale collaborations faces challenges within the current system of research evaluation. The prevailing practice of attributing credit primarily to the first author and senior (last/corresponding) author in peer-reviewed journal papers hampers the recognition of multiple contributors. This system often overlooks the valuable input of collaborators who may not fit into the traditional authorship structure but have made significant intellectual and practical contributions to the research. To truly embrace the principles of collaboration and inclusivity, there is a need for a reevaluation of the research evaluation system, allowing for proper acknowledgment of the diverse voices and contributions involved in large-scale collaborations [@ouzman2023]. Moreover, considering the checkered disciplinary history of archaeology/anthropology featuring colonial exploitation, the changes in the evaluation system alone are not enough. This further highlights the need for adopting a community-based approach to fundamentally transform the power dynamics in archaeological knowledge production and distribution [@schneider2020; @lasalle2010; @atalay2012].

@@

```{r behatrix, echo=FALSE, out.width = '150%', fig.align='center', fig.cap="The user interface of Behatrix displaying the transition probability between each action category in a handaxe knapping session."}
knitr::include_graphics(here::here("figure/Fig3.jpg"))
```

@@

Direct applications of ethnographic research in experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including contemporary U.S. hobbyists [@whittaker2004] and knapping practitioners in various non-industrial societies [@stout2002; @arthur2018], are far from novel. However, ethnography has never been formally recognized as a legitimate research method in mainstream experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information (e.g., intention, phenomenology) that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. Within the broader context of burgeoning interest in mixed-method research in contemporary social science [@creswell2017], this also echoes the post-positivist turn in psychology in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].

Through participant observation, interviews, and detailed field notes, ethnography can capture the subtle nuances of perception, such as cognitive affordances [@hussain2021; @roepstorff2008], sensory experiences [@day2013; @oneill2019; @therout2019], social interactions, and cultural meanings associated with the experimental activities [@gowlland2019]. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in the experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage. For example, serial formation of step-fracture on the debitage surface is commonly interpreted as unintentional mistakes indicative of novice knappers, while in some cases researchers treat it as evidence of deliberate core rejuvenation [@akerman1993: 126]. The accumulation of testimonies by participating knappers in terms of their intended outcome becomes useful in this scenario, although these materials should be examined in combination with the relevant product- and process-level data in a careful manner. Instead of seeing intention as something abstruse or unapproachable in archaeology [@david2004; @russell2004], the Triple P framework adopts a novel definition proposed by Quillien and German [-@quillien2021: 1] from the perspective of causal perception, namely "an agent did X *intentionally* to the extent that X was causally dependent on how much the agent wanted X to happen (or not to happen)." In this sense, the mismatch between how different individuals perceive cause-and-effect relationships and how they are organized according to physical laws is exactly where interesting variation emerges and where ethnography becomes necessary.

## Multi-level sample and data curation

@@

Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016; @nicholson2023] and CARE [@carroll2020; @gupta2023], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project. Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researchers; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.

On top of the digital data curation, an easily ignored but crucial aspect regarding the integrity and reliability of research in experimental archaeology is the long-term and proper curation of psychical specimens produced during experiments, which is particularly relevant to the product-level data. Haythron et al. [-@haythorn2018] sharply pointed out that re-running statistical analyses on a publicly available spreadsheet containing incorrect lithic projectile attribute data would be meaningless. In this case, a reexamination of the actual experiment samples becomes necessary. Moreover, it is likely that new research questions can only be answered through direct observation and measurement of the experimental assemblages themselves instead of data readily available in previously compiled spreadsheets [@eren2016]. The existence of these possibilities thus requires that an experimental assemblage of interest should be curated in public institutions with easy access and detailed contextual information [@haythorn2018].

@@

I thank Dietrich Stout for his helpful comments on earlier drafts of this article as well as Mark Moore and Shaya Jannati for inspiring discussions. I am also grateful to Metin Eren and two other anonymous reviewers for their insightful feedback. I would also like to express my sincere gratitude to Javier Baena Preysler, who helped proofread the Spanish translation of abstract and keywords.

# Funding Statement

This study was supported by the Leakey Foundation research grant titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" as well as the International Society of Human Ethology's Owen Aldis Award project titled "Sealed in stones: The computational ethology of stone toolmaking and its implications to the evolution of cultural transmission."

# Data Availability Statement

No original data were used.

# Competing Interests

The author declare none.

# Notes

1.  Strategy 2 is particularly relevant here, which is "the pursuit of general questions in present-day material culture in order to acquiring laws for making useful behavioral inferences" [@schiffer2010: 6]
2.  It should be noted that the definitions of internal and external validity constructs here solely concern the interpretation of experimental design and thus are different from the one previously proposed by Lycett and Eren [@eren2016; @lycett2013].

# References Cited
